---
title: "LQG model: random walk"
author: "Joan López-Moliner"
format: html
editor: visual
---

Just to make sure we have the functions

```{r}
#| label: load
#| warning: false


if(!("Rcpp" %in% (.packages()))){
  require(Rcpp)
  require(RcppArmadillo)
}
require(tidyverse)

sourceCpp("../cpp/cpsyLQG.cpp")

source("../R/functions.R")

```

To illustrate the LQG (linear quadratic gaussian), we use a simplest model which is a random walk on the position of a target that we want to track. Unlike the Kalman model in which we only have the dynamics, here we also have the controler that performs the action of tracking. 

The R code here is based on the LQG code from the LQG tutorial at Constantin Rothkopt lab [LQG tutorial](https://github.com/RothkopfLab/lqg/blob/main/notebooks/Tutorial.ipynb "LQG tutorial")

For further information the reference to look at is: 

Straub, D., & Rothkopf, C. A. (2022). Putting perception into action with inverse optimal control for continuous psychophysics. [Elife, 11, e76635.](https://doi.org/10.7554/eLife.76635)


The stimulus dynamics is defined as:

$${\bf x}_{t+1}= A{\bf x}_t + B{\bf u}_t + V\epsilon_t, \ \ \epsilon_t \sim \mathcal{N}(0,I)$$
and the observation equation:
$${\bf y}_t = C{\bf x}_t + W\eta_t, \ \ \eta_t\sim \mathcal{N}(0,I)$$



It is assumed that the actor solves the LQG problem by minimizing a quadratic cost function:

$$J(u_{1:T}) =\sum_{t=1}^T {\bf x}_t^TQ{\bf x}_t + {\bf u}_t^T R{\bf u}_t  $$

We will define the corresponding matrices to solve a simple random walk process and showcase an example.

$$A=\left( 
\begin{array}{}
1 & 0 \\
0 & 1
\end{array}
\right),\   
B=\left(
\begin{array}{}
0 \\
dt 
\end{array}
\right), \ 
C = \left( 
\begin{array}{}
1 & 0 \\
0 & 1
\end{array}
\right), \ 
V=\left( 
\begin{array}{}
\sigma_{rw} & 0 \\
0 & \sigma_{m}
\end{array}
\right), \ 
W=\left( 
\begin{array}{}
\sigma_{target} & 0 \\
0 & \sigma_{cursor}
\end{array}
\right), \ 
Q = \left( 
\begin{array}{}
1 & -1 \\
-1 & 1
\end{array}
\right), \ 
R= \left(c\right)
$$

Let's have a closer look at the process dynamics. $A$ is a 2x2 matrix, which means we have two states. Unlike before in which we had position and velocity, we now have target position $x^T$ and cursor position $x^C$ 

$$
\left(
\begin{array}{ll}
x_{t+1}^T\\
x_{t+1}^C
\end{array}
\right) = 
\Biggl(
\begin{array}{ll}
1 & 0\\
0 & 1
\end{array}
\Biggr) \cdot
\left(
\begin{array}{}
x_{t}^T\\
x_{t}^C
\end{array}
\right) + 
\Biggl(
\begin{array}{ll}
\sigma_{target} & 0 \\
0 & \sigma_{cursor}
\end{array}
\Biggr ) \cdot \eta_t, \ \eta_t\sim \mathcal{N}(0,I)
$$


The observation uncertainty is expressed as a two states matrix $W$, the uncertainty of perceiving the target ($\sigma_{target}$) and the uncertainty of perceiving the cursor $\sigma_{cursor}$:

$$
\left(
\begin{array}{ll}
y_{t}^T\\
y_{t}^C
\end{array}
\right) = 
\Biggl(
\begin{array}{ll}
1 & 0\\
0 & 1
\end{array}
\Biggr) \cdot
\left(
\begin{array}{}
x_{t}^T\\
x_{t}^C
\end{array}
\right) + 
\Biggl(
\begin{array}{}
0 \\ dt \\
\end{array}
\Biggr ) \cdot u_t + 
\Biggl(
\begin{array}{ll}
\sigma_{rw} & 0 \\
0 & \sigma_m
\end{array}
\Biggr ) \cdot \eta_t, \ \eta_t\sim \mathcal{N}(0,I)
$$

## A simple example to illustrate how to compute the cost

Suppose a cursor is tracking a target, so the quadratic cost function is the squared difference between the position of the target $x_1$ and that of the cursor $x_2$. The controller $Bu_t$ will move so that minimizes this cost function 
$$
\begin{pmatrix} x_1 & x_2 \end{pmatrix}
\begin{pmatrix} 1 & -1 \\ -1 & 1 \end{pmatrix}
\begin{pmatrix} x_1 \\ x_2 \end{pmatrix}
= x_1^2 - 2x_1x_2 + x_2^2 = (x_1 - x_2)^2
$$
One first multiply $Q{\bf x}_t$ and then multiply $\mathbf{x}_t^\top (Q \mathbf{x}_t)$. Of course the program does it for you. You only need to specify the right size matrices.

$$
\mathbf{x}_t =
\begin{pmatrix}
2 \\
3
\end{pmatrix}, \quad
Q =
\begin{pmatrix}
1 & -1 \\
-1 & 1
\end{pmatrix}
$$

So suppose that at time $t$ the target is at position 2 and the cursor is at position 3,
$$
\mathbf{x}_t^\top Q \mathbf{x}_t =
\begin{pmatrix} 2 & 3 \end{pmatrix}
\begin{pmatrix} 1 & -1 \\ -1 & 1 \end{pmatrix}
\begin{pmatrix} 2 \\ 3 \end{pmatrix}
= \begin{pmatrix} 2 \\ & 3 \end{pmatrix}
\begin{pmatrix} -1\ \ 1 \end{pmatrix}
= 2 \cdot (-1) + 3 \cdot 1 = -2 + 3 = 1
$$


```{r}

# Define the state vector x_t
x_t <- matrix(c(2, 3), ncol = 1)

# Define the cost matrix Q
Q <- matrix(c(1, -1,
              -1, 1), nrow = 2, byrow = TRUE)

# Compute the quadratic form x_t^T Q x_t
cost <- t(x_t) %*% Q %*% x_t

# Print the result
print(cost)
```

## Model definition

We assume we have a bounded actor (Straub & Rothkopt (2022) eLife) in which we also minimize an action cost R. 
We will define a model for the process dynamics and another for the actor.

```{r}
#| label: define matrices of the dynamics and actor

#bounded actor


action_variability = .5 #0.5
pos_noise = 1
sigma_target = 6  #6 # target perceptual variab
sigma_cursor = 1 #1.0 # cursor prop+visual variab
action_cost = .05
dt = 1. / 60.

A <- diag(1,2,2)
B <- matrix(c(0,dt),2,1)
C <- diag(1.0,2,2)
V <- diag(c(pos_noise,action_variability),2,2)
W <- diag(c(sigma_target,sigma_cursor),2,2)
dyn <- list(A=A,B=B,C=C,V=V,W=W)
Q <- matrix(c(1,-1,-1,1),2,2)
R <- matrix(action_cost,1,1)
actor <- list(A=A,B=B,C=C,V=V,W=W,Q=Q,R=R)


```

## Simulation of sequences

Now we can simulate some sequences:

```{r}
T <- 500

res <- lqg_simSequence(dyn,actor,matrix(c(0,0),1,2),matrix(c(0,0),1,2),T,return_all = TRUE) 
res2 <- data.frame(T=1:T,target=res[,1],cursor=res[,2])
ggplot(res2) + 
  geom_line(aes(T,target,col="target")) +
  geom_line(aes(T,cursor,col="cursor")) +
  scale_color_manual(values=c("blue","red"))

```

## Fitting of parameter

We try now to fit one parameter, for example the perceptual noise of the target

```{r}


set.seed(123)  # Set the random seed to 123
Ntrials <- 20

whol <- replicate(Ntrials, lqg_simSequence(dyn,actor,matrix(c(0,0),1,2),matrix(c(0,0),1,2),T,return_all = TRUE)) # we only return the state (not the belief)
xx <- simplify2array(whol)
xx <- aperm(xx,c(1,3,2))

optimise(lqgRW.op,c(0,10),x=xx[,,-c(3,4,5)],dyn=dyn,actor=actor) # we only need to pass the state not the estimates

```

```{r}

Ntrials <- 1

whol <- replicate(Ntrials, lqg_simSequence(dyn,actor,matrix(c(0,0),1,2),matrix(c(0,0),1,2),T,return_all = TRUE) )
xx <- simplify2array(whol)
xx <- aperm(xx,c(1,3,2))
k <- 1 

res <- data.frame(T=1:T,target=xx[,k,1],cursor=xx[,k,2],target_hat=xx[,k,3],cursor_hat=xx[,k,4])

ggplot(res) + 
  geom_line(aes(T,target,col="target")) +
  geom_line(aes(T,cursor,col="cursor")) +
  geom_line(aes(T,target_hat,col="target_estimate")) +
  geom_line(aes(T,cursor_hat,col="cursor_estimate")) +
  scale_color_manual(values=c("blue","orange", "green", "red"))

```
