---
title: "Kalman modeling: pos and vel model"
author: "Joan López-Moliner"
format: html
editor: visual
---


Just to make sure we have the functions 

```{r}

if(!("Rcpp" %in% (.packages()))){
  require(Rcpp)
  require(RcppArmadillo)
}
require(tidyverse)

sourceCpp("../cpp/cpsyKalman.cpp")

source("../R/functions.R")

```

Unlike before, now we have a two states space model: position and speed. We will simulate a uniform motion (i.e. constant speed), so the stimulus dynamics is defined as follows:

$$
\left(
\begin{array}{ll}
x_{t+1}\\
v_{t+1}
\end{array}
\right) = 
\Biggl(
\begin{array}{ll}
1 & \Delta t\\
0 & 1
\end{array}
\Biggr) \cdot
\left(
\begin{array}{}
x_{t}\\
v_{t}
\end{array}
\right)
$$

The observation matrix is like before:

$$\mathbf{y}_{t} = H\mathbf{x}_t + v_t,\ \  p(v_t) \sim \mathcal N(0,R)$$
But the uncertainty is also a two states matrix, one uncertainty for position and another for speed:

$$R=\left( 
\begin{array}{}
\sigma^2_p & 0 \\
0 & \sigma^2_v
\end{array}
\right)
$$


Note that the matrices are different to campture the two states and the dynamic. We assume we have an ideal observer, so the observation matrix $C$ is consistent with the observer to sense the position and velocity with some sensory uncertainty.

```{r}
#| label: model_defs


dt <- 1/90 # sampling rate
A <- matrix(c(1,0,dt,1),2,2)  # Just one component
C <- diag(1,2,2) # 
#kf = kfOscillation(q_pos=0.01,q_vel=0.001,sigma_p=0.05,sigma_v=0.022,dt=1/90,freq=0.1)
V = diag(c(0.01,0.00001),2,2) # process noise
W = diag(c(0.02,0.001),2,2) # observation noise 

```

Now we can simulate some sequences:

```{r}

N <- 1500

s1 <- simulSequence(A,C,V,W,x0 = c(0.0,2),xhat0 = c(0,2),T=N)

dd <- data.frame(time=1:N,x=s1[,1],xhat=s1[,2],v=s1[,3],vhat=s1[,4])

ggplot(dd) +
  geom_line(aes(time,xhat,col="Estimate"),linewidth=1.5,alpha=0.5) + # the process or state
  geom_line(aes(time,x,col="Actual")) +
  scale_color_manual("State",values=c("black","red")) +
  ylab("Position")

ggplot(dd) +
  geom_line(aes(time,v,col="Estimate")) + # the process or state
  geom_line(aes(time,vhat,col="Actual")) +
  scale_color_manual("State",values=c("black","red")) +
  ylab("Speed")


```


We try to estimate the sensory uncertainty.
Since we estimate 2 parameters, we will use optim rather than optimize.

```{r}

dim(s1) <- c(N,1,4) # N sample, 1 trial, 2=1 state + 1 estimate
optim(par=c(0.01,0.010),PosVel.op,x=s1)
```

We now replicate many trials and do the fit

```{r}

Ntrials <- 20
W = diag(c(0.02,0.001),2,2) #

whol <- replicate(Ntrials, simulSequence(A, C, V, W, x0 = c(0.0,2.0), xhat0 = c(0,2),T=1500) )
xx <- simplify2array(whol)
xx <- aperm(xx,c(1,3,2))

optim(par=c(0.01,0.010),PosVel.op,x=xx)

```


